A Classical Density-Functional Theory for Describing Water Interfaces

### 2013-07-08 papers/water-saft/paper.tex
Added and improved water-saft plots.  Fixed truncation bug in hughes-single-rod Tested everything on node10.

@@

>>>>

\fixme{Some stuff about water}.  An important aspects of water is its
ability to form hydrogen bonds.  Each water molecule contains two
hydrogen bond donor sites and two acceptor sites.

Wertheim's thermodynamic perturbation theory \fixme{expand/correct this}

In a recent paper \cite{hughes2013classical}, Hughes \emph{et al}
constructed a classical DFT for water which uses SAFT as a basis for
their free energy functional.  Hughes employed Clark \emph{et al}'s
\cite{clark2006developing} five fitting parameters for an optimal cDFT
for water and, in addition, introduces a parameter for the length
scale over which the dispersion interaction is correlated.  Clark
found that the four association site model works best to describe
water, which is physically accurate since each water molecule has four
hydrogen bonding sites.

This functional was successful in \fixme{list stuff}. However, the
correlation function for hard spheres from Yu and
Wu\cite{yu2002fmt-dft-inhomogeneous-associating} that is used in the
association free energy term was \fixme{untested}.

Schulte \emph{et al}\cite{schulte2012using} derived two approximate
expressions for the correlation function at contact for and
inhomogeneous hard sphere fluid based on the White Bear version of
fundamental measure theory (FMT). One of Schulte's expression found
the average number of spheres touching a point $\rr$ and is referred
to as a ``symmetric'' correlation functoin, $g_\sigma^S$ . The other
expression sets $\rr$ at the center of a sphere and find the average
spheres touching the solvent accessable surface, which could be
considered an ``asymmetric'' correlation function, $g_{\sigma}^A$. The
asymmetric function agreed closely with Monte Carlo simulation, and
much better than Yu and Wu's.

\section{Method}

We modify Hughes \emph{et al}'s method of creating a classical DFT
based on SAFT using Clark's four site model with five fitting
parameters and the added fitting parameter for the correlation
distance of the dispersion interaction.  We modify Hughes's
association free energy term to use Schulte's $g_{\sigma}^A$ rather
than Yu and Wu's.  The association free energy for our four-site model
has the form \fixme{I copy/pasted and modified these equations, but
  they still require a number of other equations to come before.
  Maybe only show the $g_\sigma^A$?}
\begin{align}
  F_\text{assoc}[n] &= 4 k_BT \int n_0(\xx) \zeta(\xx)
  \left(\ln X(\xx) - \frac{X(\xx)}{2} + \frac12\right) d\xx,
\end{align}
where the factor of $4$ comes from the four association sites per
molecule, the functional $X$ is the fraction of association sites
\emph{not} hydrogen-bonded, and $\zeta(\xx)$ is a dimensionless
measure of the density inhomogeneity.
\begin{align}
  \zeta(\xx) &= 1 - \frac{\mathbf{n}_{V2}\cdot\mathbf{n}_{V2}}{n_2^2}.
\end{align}
%
The fraction $X$ is determined by the quadratic equation
\begin{align}
  X(\xx) &= \frac{\sqrt{1 + 8n_0(\xx)\zeta(\xx)
      \Delta(\xx)} - 1}
  {4 n_0(\xx)\zeta(\xx)
    \Delta(\xx)}, \label{eq:X}
\end{align}a
where the functional $\Delta$ is a measure of hydrogen-bonding
probability, given by
\begin{align}
  \Delta(\xx) &= \kappaassoc g^\textit{SW}_\sigma(\xx)
  \left(e^{-\beta\epsilonassoc} - 1\right) \\
  g^\textit{SW}_\sigma(\xx) &= g^\textit{A}_\sigma(\xx) +
  \frac{1}{4}\beta\left(\frac{\partial a_1}{\partial \etadisp(\xx)} -
  \frac{\lambdadisp}{3 \etadisp}\frac{\partial a_1}{\partial \lambdadisp}\right)\label{eq:gSW},
\end{align}
% This is equation 77 of gil-villegas-1997-SAFT-VR
where $g^\textit{SW}_\sigma$ is the correlation function evaluated at
contact for a hard-sphere fluid with a square-well dispersion
potential, and $a_1$ and $a_2$ are the two terms in the dispersion
free energy.  The correlation function $g^\textit{SW}_\sigma$ is
written as a perturbative correction to the hard-sphere fluid
correlation function $g^\textit{A}_\sigma$, for which we use the
functional of Schulte \emph{et al}\cite{schulte2012using}:
\begin{align}
  g_\sigma^{A}(\rr) &= \frac{1}{n(\rr)n_A(\rr)}\frac{1}{k_B T 4 \pi
    (2R)^2}\frac{\delta F_{hs}}{\delta R(\rr)} 
\end{align}
where the $n_A(\rr)$ is the density of spheres that are touching a
sphere that is located at point $\rr$. The modification to Hughes's
functional appers here where we have replaced $g_\sigma^{HS}$ with
$g_\sigma^A$. \fixme{The equation for $g_\sigma^A$ requires finding
  convolutions of local derivative of the free energy, aking this
  formulation computationally more expensive than the free energy
  itself.}

We use Clark's five empirical parameters, and fit the calculated
surface tension to experimental surface tension at ambient conditions.
We find these agree when $\lscale$ is 0.454.

\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{figs/density-compare}
\end{center}
\caption{ Density profiles for single rods of different
  radii for the new functional (top) and the old functional
  (bottom). Each separately colored line represents a density profile
  for a rod; the black dotted line is the bulk density.}
\label{fig:density-single-rod}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{figs/single-rod-X-plot}
\end{center}
\caption{ The average number of hydrogen bonds per molecule at
  distances from the center of hydrophobic rods of various
  radii. The solid lines are the results from using the assymetric
  $g_{\sigma}^A$, while the dotted lines are from Yu and Wu's $g_{\sigma}^{HS}$.}
\label{fig:single-rod-X}
\end{figure}

\section{Hydrogen bonding near a hydrophobic rod}

We first simulate a single hydrophobic rod immersed in water. The
density profile of the water with our functional is compared with that
of Hughes in Fig. \ref{fig:density-single-rod}.  For small rods, the
water densities are attenuated near the rods for our functional as
compared to Hughes's.   Since we have changed the association term in
the functional, a noticeable change might occur in the aveerage number
of hydrogen bonds in a given system.  We compare the hydrogen bonds
with the new and old functional for rods of a few different radii
(Fig. \ref{fig:single-rod-X}) and find that there is a very noticeable
difference.  This is not suprising since $g_\sigma^A$ is antisymmetric
and $g\sigma^{HS}$ is symmetric.  \fixme{Physical description for why
  it looks that way}

\section{Hydrogen bonding near hard-sphere solutes}

\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{figs/spheres-X-plot}
\end{center}
\caption{ The average number of hydrogen bonds per molecule at
  distances from the center of spherical cavities of various
  radii. The solid lines are the results from using the assymetric
  $g_{\sigma}^A$, while the dotted lines are from Yu and Wu's $g_{\sigma}$.}
\label{fig:spheres-X}
\end{figure}

\fixme{More stuff here for spheres}

\clearpage

\section{Introduction}

@@

%% \begin{figure}
%% \begin{center}
%% \includegraphics[width=3.5in]{figs/surface-tension}
%% \end{center}
%% \caption{Comparison of Surface tension versus temperature for theoretical and
%%   experimental data. The experimental data is taken from NIST.\cite{nistwater}
%%   The length-scaling parameter $\lscale$ is fit so that the theoretical surface 
%%   tension will match the experimental surface tension near room temperature.}
%% \label{fig:surface-tension}
%% \end{figure}
